Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-03-13 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
n

The goal of this chapter is to utilize an analytical tool to predict or simulate the results from some of the key experiments conducted in this study. Unlike the previous chapter the computational fluid dynamics (CFD) models are not as widely used by fire investigators as the algebraic tools are.  The CFD models require more knowledge, more computer resources, and more input data to utilize correctly.  Currently modelers need to develop there own data or rely on poorly defined databases with the hope that the material tested in the database is similar to the specific material of interest.

@@

This seemed to be counter intuitive to the results expected from an increased RI.  However in each of the FDS models, the results closest to the burner seem to indicate that the model may not be `burning' the appropriate amount of fuel (fuel rich), which resulted in temperatures significantly lower than measured.  The affect of this fuel rich area above the burner results in poorer agreement with increasing RI.  It appears that the the RI=10 and RI=20 models follow the correct trend, however the combuston zone was moved to a higher elevation relative to what was measured.  For example the values at 0.5~m and 0.8~m above the burner, from the RI=10 and RI=20 cases respectively, seemed to coincide with the measured value at 0.3~m above the burner.  Making the adjustment appears to move the remaining temperatures above those points into alignment with the measurements along the centerline of the plume. As FDS is a complex model, this explaination may be only one possibility.

\begin{figure}[h]
	\includegraphics[width=5in]{../Figures/FHNG01_RI} \\
	\caption[Comparison of vertical centerline temperatures above the natural gas burner with FDS simulated gas temperatures.]{Comparison of vertical centerline temperatures above the natural gas burner with FDS simulated gas temperatures generated with three different RIs 5, 10, and 20.}
	\label{FHNG01_RI}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=5in] {../Figures/McCaffrey_Plume_comp}  \\
	\caption[Comparison of vertical centerline temperatures above a 0.3~m natural gas burner from McCaffrey with FDS simulated gas temperatures.] {Comparison of vertical centerline temperatures above a 0.3~m natural gas burner from McCaffrey with FDS simulated gas temperatures generated with three different RIs 5, 10, and 20 from the FDS Validation Guide~\cite{FDS_Validation_Guide}.}
	\label{McCaffrey_Plume_comp}
\end{figure}

The comparisons above used the predicted gas temperature from the model. Another comparison can be made examining the `thermocouple' results from FDS to see if the model of the thermocouple provides improved argeement with the temperatures measured via physical thermocouples.  This comparison is shown in Figure~\ref{FHNG01_RI_TC}.  Not surprisely, the sub model of the thermocouple device shifted the FDS values slightly.  At temperatures greater than 700~$^\circ$C, the `thermocouple' temperatures tended to be lower than the gas temperatures, perhaps due to radiant heat transfer losses. At temperatures less than 700~$^\circ$C, the `thermocouple' temperatures tended to be a higher than the gas temperatures, perhaps due to gains from radiant heat transfer.  The best agreement along the vertical array of temperatures was provided by the RI=5 model.  Variables such as density and emissivity required for modeling the thermal response and heat transfer to the thermocouple bead are generating the diffrences between the gas temperature and the measured temperature FDS values.

@@

The computational domain of the model was the same as the volume considered for the source fire simulations, a cube with 2.4~m on a side. The models were run with two different sized computational grid cells, an 80~mm (RI=5) grid and a 40~mm (RI=10) grid.  The face of the wall is located 1.95~m from the front of the domain.  
The wall was modeled with material properties for gypsum wallboard that were documented in Chapter 4.  The input properties were: thermal conductivity (0.28 W/m K), specific heat (1.00 kJ/kg K) and density (807 kg/m$^3$).  In addition FDS includes a default value of 0.9 for emissivity, this value should be representative of the painted surface of the gypsum wallboard. As noted earlier in this study, gypsum wallboard undergoes chemical and physical changes as it is heated.  These changes include the loss of homogeneity of the gypsum material, a change in density and as the paper is charred or burned there is a potential for a change in emissivity.  The thermal conductivity and the specific heat of the gypsum wallboard varies on the order of 20~$\%$ over the temperature range from 20 to 500~$^\circ$C~\cite{Gross:1985}.  While FDS can accomodate changes in thermoconductivy and specific heat with respect to time, density and emissivity must be held constant~\cite{FDS_Users_Guide}.  For the purposes of this study, the input values will remain constant throught out the course of the simulations.  The only variable will be the distance between the edge of the burner and the wall.

As before the FDS simultions will include devices, which are used to simulate thermocouples and total heat flux gauges.  This devices were positioned in locations similar to the locations used in the physical experiments, above the centerline of the burner and on the wall as shown in Chapter 4, Figure~\ref{Instrumented_GB_Wall}. In addition, HRRPUL will again be used as a means to gain insight into the peak flame height simluated by FDS.

@@

 There seem to be a couple of items that stick out from the graphs.  First is what appears to be a separation of the temperature data into three distinct sections.  The bottom three positions, 0.2~m through 0.6~m, represent the hottest section.  The next two measurement positions at 0.8~m and 1.0~m are similar in temperature.  The third section extends from 1.2 through 2.2~m above the burner and has the shape of a smooth curve decreasing in temperature with increasing elevation.  Comparing the temperatures to the flame height analysis in Chapter 2, the first section aligns with the median flame height, the second section aligns with the portion of the flame that intermittent, 40~$\%$ of the time or less.  The upper most section of the temperature data aligns with the thermal plume above the flame.

@@

@@

\begin{figure}[ht!]
  \centering
  \includegraphics[width=2.5in]{../Figures/IWGB_NG_HF_Center_Avg_2D}
  \includegraphics[width=2.5in]{../Figures/IWGB_NG_HF_Center_Avg_1D}\\
  \includegraphics[width=2.5in]{../Figures/IWGB_NG_HF_Center_Avg_0p5D}
  \includegraphics[width=2.5in]{../Figures/IWGB_NG_HF_Center_Avg_0D}\\
  \caption[FDS values based on two different RI values compared with heat flux measurements on the wall along the centerline of the natural gas fueled burner.]{FDS values based on two different RI values compared with heat flux measurements on the wall along the centerline of the natural gas fueled burner. The heat fluxes are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall.  Beginning with the top left graph position and viewing left to right the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.}
  \label{FDS_HFCL_IWNG_comp}
\end{figure}

\begin{figure}[ht!]
  \centering
  \includegraphics[width=2.5in]{../Figures/IWGB_NG_HF_Offset_Avg_2D}
  \includegraphics[width=2.5in]{../Figures/IWGB_NG_HF_Offset_Avg_1D}\\
  \includegraphics[width=2.5in]{../Figures/IWGB_NG_HF_Offset_Avg_0p5D}
  \includegraphics[width=2.5in]{../Figures/IWGB_NG_HF_Offset_Avg_0D}\\
  \caption[FDS values based on two different RI values compared with heat flux measurements on the wall along the right edge of the natural gas fueled burner]{FDS values based on two different RI values compared with heat flux measurements on the wall along the right edge of the natural gas fueled burner. The heat fluxes are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall.  Beginning with the top left graph position and viewing left to right the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.}
  \label{FDS_HFEdge_IWNG_comp}
\end{figure}

@@

Table to compare measured flame height with FDS cal~\cite{McDermott:2011}

Flame Height 2D  20~mm  at 99~\%  1.28~m

Flame Height 0D  20~mm  at 99~\%  1.46~m

>>>>

The goal of this chapter is to utilize an analytical tool to predict or simulate the results from some of the key experiments conducted in this study. Unlike the previous chapter the computational fluid dynamics (CFD) models are not as widely used by fire investigators as the algebraic tools are.  The CFD models require more knowledge, more computer resources, and more input data to utilize correctly.  Currently modelers need to develop their own data or rely on poorly defined databases with the hope that the material tested in the database is similar to the specific material of interest.

@@

\begin{figure}[h!]
	\includegraphics[width=5in]{../Figures/FHNG01_RI} \\
	\caption[Comparison of vertical centerline temperatures above the natural gas burner with FDS simulated gas temperatures.]{Comparison of vertical centerline temperatures above the natural gas burner with FDS simulated gas temperatures generated with three different RIs 5, 10, and 20.}
	\label{FHNG01_RI}
\end{figure}

This seemed to be counter intuitive to the results expected from an increased RI.  However in each of the FDS models, the results closest to the burner seem to indicate that the model may not be `burning' the appropriate amount of fuel (fuel rich), which resulted in temperatures significantly lower than measured.  The effect of this fuel rich area above the burner results in poorer agreement with increasing RI.  It appears that the the RI=10 and RI=20 models follow the correct trend, however the combuston zone was moved to a higher elevation relative to what was measured.  For example the values at 0.5~m and 0.8~m above the burner, from the RI=10 and RI=20 cases respectively, seemed to coincide with the measured value at 0.3~m above the burner.  Making the adjustment appears to move the remaining temperatures above those points into alignment with the measurements along the centerline of the plume. As FDS is a complex model, this explaination may be only one possibility.

@@

\begin{figure}[h]
	\includegraphics[width=5in] {../Figures/McCaffrey_Plume_comp}  \\
	\caption[Comparison of vertical centerline temperatures above a 0.3~m natural gas burner from McCaffrey with FDS simulated gas temperatures.] {Comparison of vertical centerline temperatures above a 0.3~m natural gas burner from McCaffrey with FDS simulated gas temperatures generated with three different RIs 5, 10, and 20 from the FDS Validation Guide~\cite{FDS_Validation_Guide}.}
	\label{McCaffrey_Plume_comp}
\end{figure}

The comparisons above used the predicted gas temperature from the model. Another comparison can be made examining the `thermocouple' results from FDS to see if the model of the thermocouple provides improved argeement with the temperatures measured via physical thermocouples.  This comparison is shown in Figure~\ref{FHNG01_RI_TC}.  Not surprisingly, the sub model of the thermocouple device shifted the FDS values slightly.  At temperatures greater than 700~$^\circ$C, the `thermocouple' temperatures tended to be lower than the gas temperatures, perhaps due to radiant heat transfer losses. At temperatures less than 700~$^\circ$C, the `thermocouple' temperatures tended to be a higher than the gas temperatures, perhaps due to gains from radiant heat transfer.  The best agreement along the vertical array of temperatures was provided by the RI=5 model.  Variables such as density and emissivity required for modeling the thermal response and heat transfer to the thermocouple bead are generating the diffrences between the gas temperature and the measured temperature FDS values.

@@

The computational domain of the model was the same as the volume considered for the source fire simulations, a cube with 2.4~m on a side. The models were run with two different sized computational grid cells, an 80~mm (RI=5) grid and a 40~mm (RI=10) grid.  The face of the wall is located 1.95~m from the front of the domain.  
The wall was modeled with material properties for gypsum wallboard that were documented in Chapter 4.  The input properties were: thermal conductivity (0.28 W/m K), specific heat (1.00 kJ/kg K) and density (807 kg/m$^3$).  In addition FDS includes a default value of 0.9 for emissivity, this value should be representative of the painted surface of the gypsum wallboard. As noted earlier in this study, gypsum wallboard undergoes chemical and physical changes as it is heated.  These changes include the loss of homogeneity of the gypsum material, a change in density and as the paper is charred or burned there is a potential for a change in emissivity.  The thermal conductivity and the specific heat of the gypsum wallboard varies on the order of 20~$\%$ over the temperature range from 20 to 500~$^\circ$C~\cite{Gross:1985}.  While FDS can accomodate changes in thermoconductivy and specific heat with respect to time, density and emissivity must be held constant~\cite{FDS_Users_Guide}.  For the purposes of this study, the input values will remain constant throughout the course of the simulations.  The only variable will be the distance between the edge of the burner and the wall.

As before, the FDS simultions will include devices which are used to simulate thermocouples and total heat flux gauges.  These devices were positioned in locations similar to the locations used in the physical experiments, above the centerline of the burner and on the wall as shown in Chapter 4, Figure~\ref{Instrumented_GB_Wall}. In addition, HRRPUL will again be used as a means to gain insight into the peak flame height simluated by FDS.

@@

 There seem to be a couple of items that stick out from the graphs.  First is what appears to be a separation of the temperature data into three distinct sections.  The bottom three positions, 0.2~m through 0.6~m, represent the hottest section.  The next two measurement positions at 0.8~m and 1.0~m are similar in temperature.  The third section extends from 1.2 through 2.2~m above the burner and has the shape of a smooth curve decreasing in temperature with increasing elevation.  Comparing the temperatures to the flame height analysis in Chapter 2, the first section aligns with the median flame height, the second section aligns with the portion of the flame that is intermittent, 40~$\%$ of the time or less.  The upper most section of the temperature data aligns with the thermal plume above the flame.

@@

\begin{figure}[ht!]
  \centering
  \includegraphics[width=2.5in]{../Figures/IWGB_NG_HF_Center_Avg_2D_RI}
  \includegraphics[width=2.5in]{../Figures/IWGB_NG_HF_Center_Avg_1D_RI}\\
  \includegraphics[width=2.5in]{../Figures/IWGB_NG_HF_Center_Avg_0p5D_RI}
  \includegraphics[width=2.5in]{../Figures/IWGB_NG_HF_Center_Avg_0D_RI}\\
  \caption[FDS values based on two different RI values compared with heat flux measurements on the wall along the centerline of the natural gas fueled burner.]{FDS values based on two different RI values compared with heat flux measurements on the wall along the centerline of the natural gas fueled burner. The heat fluxes are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall.  Beginning with the top left graph position and viewing left to right the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.}
  \label{FDS_HFCL_IWNG_comp}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
  \includegraphics[width=2.5in]{../Figures/IWGB_NG_HF_Offset_Avg_2D_RI}
  \includegraphics[width=2.5in]{../Figures/IWGB_NG_HF_Offset_Avg_1D_RI}\\
  \includegraphics[width=2.5in]{../Figures/IWGB_NG_HF_Offset_Avg_0p5D_RI}
  \includegraphics[width=2.5in]{../Figures/IWGB_NG_HF_Offset_Avg_0D_RI}\\
  \caption[FDS values based on two different RI values compared with heat flux measurements on the wall along the right edge of the natural gas fueled burner]{FDS values based on two different RI values compared with heat flux measurements on the wall along the right edge of the natural gas fueled burner. The heat fluxes are averages of two replicate experiments. Each graph displays data from a different distance between the edge of the burner and the instrumented wall.  Beginning with the top left graph position and viewing left to right the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.}
  \label{FDS_HFEdge_IWNG_comp}
\end{figure}

@@

Complete to compare measured flame height with FDS cal~\cite{McDermott:2011}

\begin{table}[h]
  \small
  \centering
  \begin{tabular}{|c|c|c|c|} \hline 
Distance from Wall 		&  Flame Height 			  	& RI=5       		& RI=10      		  \\
    					&  Range          				&  (80~mm grid)  	&  (40~mm grid)  	  \\
    (m)					&  (m) 							& (m) 				& (m) 				  \\ 
\hline    0.61			&  0.73 to 1.14         		& 1.02 to 1.27      & 1.xx to 1.28         \\
\hline    0.31			&								&					&					    \\
\hline    0.15			&								&					&						\\
\hline	  0.0			&								&					&  1.xx to 1.46 	  \\
\hline
  \end{tabular}
  \caption[Comparison of estimations of flame height from the natural gas instrumented wall experiments and FDS simulations with RIs of 5 and 10.]{Comparison of estimations of flame height from the natural gas instrumented wall experiments and FDS simulations with RIs of 5, and 10.}
  \label{tab:FDS_FH_IWNG}
\end{table}
